---
title: "Lecture 1 - Probability Distributions"
author: "Sam Mason"
date: "1/28/2020"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Part 1: Exploring the Shape of Continuous Distributions

The shape of a normal distribution is given by its first and second moments, the mean $\mu$ and variance $\sigma^2$. In this figure below, the first row of distributions holds $\mu$ constant at 5, and allows $\sigma^2$ to vary. The second row features normal distributions with constant $\sigma^2$ and variable $\mu$. By definition, the normal distribution has skewness and kurtosis equal to 0.

```{r normal distributions, echo=FALSE}
par(mfrow=c(2, 3))
hist(rnorm(1000, 5, 0.5), breaks = 50, main = 'Variance = 0.5', xlim = c(-10, 20), freq = F, xlab = NULL, col = 'black')
hist(rnorm(1000, 5, 2), breaks = 50, main = 'Variance = 2', xlim = c(-10, 20), freq = F, xlab = NULL, col = 'black')
hist(rnorm(1000, 5, 5), breaks = 50, main = 'Variance = 5', xlim = c(-10, 20), freq = F, xlab = NULL, col = 'black')
hist(rnorm(1000, 0, 1), breaks = 50, main = 'Mean = 0', xlim = c(-6, 6), freq = F, xlab = NULL, col = 'black')
hist(rnorm(1000, -3, 1), breaks = 50, main = 'Mean = -3', xlim = c(-6, 6), freq = F, xlab = NULL, col = 'black')
hist(rnorm(1000, 3, 1), breaks = 50, main = 'Mean = 3', xlim = c(-6, 6), freq = F, xlab = NULL, col = 'black')
```

The normal distribution is appropriate for continuous variables, those which can take on the values of all real numbers (of which there are infinitely many) within a given interval. We call this interval the **support**, and say things things like *the normal distribution has support in the set of all real numbers*. Let's look at another continuous distribution useful to biologists, the beta distribution with support along the interval (0, 1). Unlike the normal distribution, the beta distribution has parameters $\alpha$ and $\beta$, which do not explicitly reflect the first and second moments of the distribution. For example, the first moment of a beta distribution is given by the ratio $\frac{\alpha}{\alpha+\beta}$. Let's prove this to ourselves! I'll show you an R proof using the normal distribution, which will be mathemtically unfulfilling, but syntactically helpful.

```{r proving the mean of a normal distribution}
# First we generate a large random sample a normal distribution using the rnorm() function
samp <- rnorm(1000, 2, 3)

# Use the help call ? to determine the parameters defined by each of the three arguments.
# Let's plot this data using the hist() function to visualize it.
hist(samp, breaks = 50, freq = FALSE, col = 'black')

# What is the "breaks" argument doing here? Looks like the first moment is right where 
# it's supposed to be around two. Let's confirm by using R's mean() function. 
mean(samp)

# Pretty close! And we'll get closer for larger n. The first moment of a normal
# distribution is equal to it's mean parameter. So here's the big question: is 
# two equal to two? Let's submit our query humbly before R:
2 == 2

# R HAS SPOKEN! This is an example of a logical expression, and can be evaluated by
# R as either "TRUE" or "FALSE"
```

Alright, you're up! Here are the steps:  
  
1. Create a vector of 1000 random draws from a beta distribution with $\alpha$ and $\beta$ parameters of your choosing. **HINT:** make sure you're assigning this vector to a new object so you can use it later.
2. Calculate the mean using the ```mean()``` function.
3. Calculate the mean using the $\alpha$ and $\beta$ parameters defined in step 1.  

## Part II: Discrete Distributions and the Probability Mass Function (PMF)

Let's take a look at a discrete distribution, the Bernoulli distribution, with support in the set {0, 1}. To be clear, Bernoulli data is binomial, and can only take on values of zero and one. Can you think of an example of real-world binomial data? Perhaps the presence (one) or absence (zero) of a particular butterfly species in a sample of salt marsh patches? Let's dig into this concept of binomial data a bit more. Because the data can only exist in two states (zero or one), we say that binomial states are *disjoint*, that is, knowledege of one state implies perfect knowledge of the other. If a patch of salt marsh is occupied by the butterfly species (one), it cannot possibly be unoccupied (zero). Because a patch can only exist as occupied or unoccupied, if the probability of occupancy is $\Psi$, then the probability of unoccupancy is $1-\Psi$. Because of this, when defining the shape of a Bernoulli distribution, we need only specify the probability of one state or the other (conventionally the *probability of success*, $P(1)$).  

We can describe the shape of a Bernoulli distribution using the appropriate probability mass function (PMF) presented below:  

$$
PMF = f(k) = \begin{cases}
q = 1 - p \;\;\;\;\; if \;\; k = 0\\
p \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\, if \;\; k = 1
\end{cases}
$$
  
where $p$ is the probability of success ($k=1$) and $q$ is the probability of failure ($k=0$). The PMF, $f(k)$ takes some value $k \in {0, 1}$ and returns the probability of that value. This is important to note: the PMF of a discrete distribution calculates the *probability* of a discrete value existing in the support of the distribution. In the next section we'll introduce the probability *density* function (PDF) associated with continuous distributions, and compare it to the concept of the PMF.  

The first and second moments of the Bernoulli distribution are given by $p$ and $pq$, respectively. Let's exmaine how these two moments relate to one another. Please create a scatter plot where the first moment is on the x-axis, and the second moment is on the y-axis. Here's how we'll tackle it:  
  
1. Create a vector called "m1" and assign it all values from 0 to 1, incremented by 0.1 using the ```seq()``` function in R.
2. Create a new vector, "q" equal to $1-m1$. In R, we can apply arithmetic operators, like substraction, to each element of a vector by simply using the vector in an arithmetic expression. For example, if I wanted to add five to each element of "m1" I could execute the code ```m1 + 5```.
3. Create a third vector, "m2", equal to the product of "m1" and "q".
4. Use the plot() function with default arguments to create your scatterplot.  

Here's what my plot looks like after I spruced up the formatting a bit:

```{r plotting the moments of a Bernoulli distribution, echo=FALSE}
m1 <- seq(0, 1, 0.1)
q <- 1-m1
m2 <- m1*q
plot(m1, m2, pch=19, xlab = 'First Moment (p)', ylab = 'Second Monent (pq)')
```

## Part III: Probability Mass vs. Density

We'll start by introducing a new continuous distribution, the gamma distribution, with support $x \in (0, \infty)$, and parameters $\alpha$ (shape) and $\beta$ (rate). As with any new distribution, let's take a moment to explore it. Please make a 2x3 (row by column) matrix of gamma distributions, holding shape constant in the first row, and rate constant in the second row (similar to what I produced above for the normal distribution). The specific parameter values are up to you. We'll begin to address graphics in R more directly in this exercise. Okay, let's dive in:  

1. When composing graphics in R, we often want to set some formatting parameters ahead of time. The function ```par()``` does the heavy lifting in this regard. This is a very powerful function, but for now we'll use it to create our plotting matrix and define subplot margins. Begin your plotting with the following line: ```par(mfrow=c(2, 3), mar = c(3, 3, 3, 3))```. The "mfrow" argument says to R: *I want six subplots, organized as two rows of three plots, and I want new plots to be placed in the matrix by row (i.e., the first plot gets placed in cell (1, 1), the second plot in cell (1, 2), etc.)*. The "mar" argument defines how large the margins around each subplot should be. The margins are presented as a list in the order (*bottom, left, top, right*).
2. Create a vector containing 1000 random draws from a gamma distribution with parameters of your choosing.
3. Using the ```hist()``` function, plot your first distribution. Use the help call ? to identify the arguments that hist() can take. Some key formatting include: **"main"**, specifying the title of the plot (must put title in quotation marks); **"breaks"**, the number of bins included in your histogram; **"freq"**, a logical argument accepting either TRUE or FALSE (TRUE: plots count on the y-axis, FALSE: plots probability density); **"xlab"**, specifying the x-axis label (quotations required); **"ylab"**, specifying the y-axis label (quotations required); **"xlim"**, determines the extent of the domain (specified as a concated list, for example, ```c(0, 10)```); **"col"**, the color of the bars.  

**HINT:** in order to adequately compare distributions, make sure your "xlim" argument is consistent among relevant plots.  
**HINT:** if you mess up a subplot, you'll need to clear the graphics memory using the broom icon in the "Plots" window of RStudio, and then rerun the code, including the ```par()``` line.
